---
title: "Impact of Oil Palm Production on Forest Cover and Carbon Emissions in Malaysia and Indonesia  "
author: "Ayoung Kim, Queenie Wei, Vanshika Mittal"
date: "Dec 12, 2023"
output: pdf_document
geometry: margin=2.54cm
toc: true
---

\newpage

| Table  | Name                               |
|-----------:|:-------------------------------------|
| 1   | Palm Production                       |
| 2   | Forest Area                           |
| 3   | Carbon Emissions                      |

: List of Tables 

| Figures  | Name                             |
|-----------:|:-------------------------------------|
| 1    | Palm Production                      |
| 2    | Forest Area                          |
| 3    | Carbon Emissions                     |

: List of Figures

\newpage

```{r setup, include=FALSE}
# Set your working directory
getwd()
# Load your packages
library(tidyverse)
library(lubridate)
library(ggplot2)
library(raster)
library(here)
library(sf)
library(leaflet)
library(agricolae)
library(rvest)
library(vctrs)
library(mapview)
library(dplyr)
library(tseries)
#install.packages("forecast")
library(trend)
# Set your ggplot theme
mytheme<-theme_classic(base_size=13)+
theme(
legend.background = element_rect(color='grey',fill='beige'),
plot.title=element_text(color='black',size=15),
legend.position="right",
legend.text = element_text(size=11)
)

```

# Load your datasets
```{r}
#dataset on palm cultivation 
palm_production <-read.csv("./palm-oil-production.csv",stringsAsFactors = TRUE)
#filtering for only Indonesia and Malaysia, making a new column with an adjusted scale for oil palm cultivation 
palm_production <- palm_production %>%
  filter(Entity == "Indonesia"| Entity == "Malaysia", Year %in% (1990:2020))%>%
  mutate(Oil.Palm = Palm.oil...00000257....Production...005510....tonnes/1000) %>%
  dplyr::select(Entity, Year, Oil.Palm)
  
#dataset on forest area 
forest_area <- read.csv("./forest-area-primary-planted.csv",stringsAsFactors = TRUE)
#filtering for Indonesia and Malaysia, adjusting scales for planted and natural forests
forest_area <- forest_area %>%
  filter(Entity == "Indonesia" | Entity == "Malaysia") %>%
  mutate(Planted.Forest = 
           Planted.Forest...00006716....Area...005110....hectares/1000000 ) %>%
  mutate(Natural.Forest = 
           Naturally.regenerating.forest...00006717....Area...005110....hectares/1000000)%>%
  dplyr::select(Entity, Year, Planted.Forest, Natural.Forest)

#datasets on carbon emissions 
carbon_emissions <- read.csv("./ghg-emissions-by-sector_indonesia.csv", stringsAsFactors = TRUE)
#filtering for Indonesia and Malaysia, adjusting scales for co2 emissions variable
carbon_emissions <- carbon_emissions %>%
  filter(Entity == "Indonesia"| Entity == "Malaysia")%>%
  mutate(CO2.emissions = Greenhouse.gas.emissions.from.land.use.change.and.forestry/1000000)%>%
  dplyr::select(Entity, Year, CO2.emissions)
```

```{r}
#merge the datasets into one for analysis 
merged_dataset <- merge(palm_production, forest_area, by = c("Entity", "Year"), 
                 all = TRUE)
dataset <- merge(merged_dataset, carbon_emissions, by = c("Entity", "Year"), 
                 all = TRUE)
dataset <- dataset %>%
  mutate(Total.Forest = Planted.Forest + Natural.Forest)
```

# Rationale and Research Questions
>The oil palm tree is a productive and useful crop, as it can be cultivated at a lower price, providing high-quality oil (WWF, n.d.). It is used not only as food but for other purposes, such as in detergents/ cosmetics and for bioenergy (Ritchie, 2021). This has led to a rapid expansion in the oil palm industry, particularly across Asia, Africa, and Latin America (WWF, n.d.). For example, the yield increased from 2 million tonnes in 1970 to 71 million tonnes in 2018 (Ritchie, 2021). According to the Food and Agriculture Organization of the United Nations, Indonesia and Malaysia contributed to 57% and 27% of global production respectively (Richie, 2021). However, the expansion of the industry has been linked to some negative impacts on the environment, becoming one of the major contributors to deforestation, carbon emissions, and tropical biodiversity loss (WWF, n.d.). 
>In this project, we will explore the oil palm cultivation industry in Indonesia and Malaysia. We want to measure how the cultivation of oil palm has contributed to deforestation and carbon emissions in the two countries from 1990 to 2020.” 
>H0: The cultivation of oil pam has not significantly changed forest cover (carbon emissions) in Indonesia and Malaysia. 
>Ha: The cultivation of oil palm has significantly changed forest cover (carbon emissions) in Indonesia and Malaysia. 

\newpage

# Dataset Information
>Firstly, we utilized the 'palm_production' dataset from 'Our World In Data' to analyze the changes in the oil palm industry. This dataset includes information on the country and continent, year, and the amount of production in tonnes. Initially, we selected the entities (country names), year, and 'Palm.oil...00000257....Production...005510....tonnes', representing the production quantity. We then used the 'filter' function to narrow down the data to Indonesia and Malaysia.For the analysis of land coverage, we obtained the forest area dataset from ‘Our World in Data’. This dataset comprises six variables: country name, country code, year (date), indicator name, indicator code, and indicator value in number. We filtered the data using 'AG.LND.FRST.K2' as the indicator code and selected country name, year (date), and the indicator value in number. The carbon emissions data for both Indonesia and Malaysia, titled 'ghg-emissions-by-sector_indonesia' from Our World in Data, provides information on greenhouse gas emissions from various sectors such as agriculture, land use change and forestry, waste, buildings, industry, manufacturing and construction, transport, electricity and heat, fugitive emissions of greenhouse gases from energy production, and greenhouse gas emissions from fuel combustion and bunker fuels. We employed the 'filter' function to extract data specifically for Indonesia and Malaysia, selecting 'Entity' (country name), 'year', and 'Greenhouse.gas.emissions.from.land.use.change.and.forestry' from the 'carbon_emissions' dataset.


| Data  | Name                               | Variables of Interest            |
|-----------:|:-------------------------------------:|:--------------------------|
| Oil Palm Production  | palm-oil-production.csv  | Entity, Year, Palm.oil...00000257....Production...005510....tonnes |
| Primary Forests | forest-area-primary-planted.csv  | Entity, Year, Planted.Forest...00006716....Area...005110....hectares |
| Natural Forests | forest-area-primary-planted.csv   | Entity, Year, Naturally.regenerating.forest...00006717....Area...005110....hectares | 
| Carbon Emissions  | ghg-emissions-by-sector_indonesia.csv | Entity, Year,  Greenhouse. gas emissions.from.land.use .change.and.forestry |

\newpage

# Exploratory Analysis 

```{r}
#plotting oil palm cultivation over the years to see the trend 
options(scipen = 999)
palm_production_graph <- ggplot(palm_production, aes(x=Year, y = Oil.Palm, color = Entity)) + 
  geom_point()+ 
  geom_smooth(method = lm)+ 
  labs(x= "Years", y = "Palm Oil Production (Thousand Tonnes)", title = "Palm Production in Indonesia and Malaysia over the years")

palm_production_graph
  
#graphing the data on forest area 
forest_area_graph <- ggplot(forest_area, aes(x=Year)) +
                     geom_line(aes(y = Planted.Forest), color = "green") + 
                     geom_line(aes(y = Natural.Forest, color = "red")) + 
                    facet_wrap(vars(Entity))+
                    labs(x= "Years", y = "Forest Area (Million Hectares)", title = 
                           "Forest Area over the years")

forest_area_graph

#graphing dataset on carbon emissions 
carbon_emissions_graph <- ggplot(carbon_emissions, aes(x=Year, y = CO2.emissions, color = Entity)) +
  geom_point()+ 
  geom_smooth(method = lm)+ 
  labs(x= "Years", y = "Carbon Dioxide Emissions (Million tonnes)", title = "Carbon Dioxide Emissions over the years")

carbon_emissions_graph
```
\newpage

# Analysis
```{r, testing research question 1}
#MULTIPLE LINEAR REGRESSION
#checking best fit models through AIC 
natural_AIC <- lm(data = dataset, Natural.Forest ~ Year + Oil.Palm)
step(natural_AIC)

planted_AIC <- lm(data = dataset, Planted.Forest ~ Year + Oil.Palm)
step(planted_AIC)

total_AIC <- lm(data = dataset, Total.Forest ~ Year + Oil.Palm)
step(total_AIC)
#include both oil palm and year, they have the lowest AIC 

#Planted Forests in Indonesia 
planted_ind <- lm(data = filter(dataset, Entity == "Indonesia") , Planted.Forest ~ Oil.Palm + Year)
summary(planted_ind)

#Natural Forests in Indonesia 
natural_ind <- lm(data = filter(dataset, Entity == "Indonesia") , 
                  Total.Forest ~ Oil.Palm + Year)
summary(natural_ind)

#Total Forests in Indonesia
total_ind <- lm(data = filter(dataset, Entity == "Indonesia") ,
                  Total.Forest ~ Oil.Palm + Year)
summary(total_ind)

#Planted Forests in Malaysia
planted_mly <- lm(data = filter(dataset, Entity == "Malaysia") ,
                  Planted.Forest ~ Oil.Palm + Year)
summary(planted_mly)

#Natural Forests in Malaysia
natural_mly <- lm(data = filter(dataset, Entity == "Malaysia") , 
                  Natural.Forest ~ Oil.Palm + Year)
summary(natural_mly)

#Total Forests in Malaysia
total_mly <- lm(data = filter(dataset, Entity == "Malaysia") ,
                  Total.Forest ~ Oil.Palm + Year)
summary(total_mly)
```

```{r, testing reaseach question 2}
#checking best fit models through AIC 
co2_AIC <- lm(data = dataset, CO2.emissions ~ Year + Oil.Palm + Natural.Forest + Total.Forest + Planted.Forest)
step(co2_AIC) 
#model to be used is with year, oil palm and total forests - has the lowest AIC 

#MLR checking for CO2 in Indonesia 
CO2_ind <- lm(data = filter(dataset, Entity == "Indonesia") , 
                  CO2.emissions ~ Total.Forest + Oil.Palm + Year)
summary(CO2_ind)

#MLR checking for CO2 in Malaysia
CO2_mly <- lm(data = filter(dataset, Entity == "Malaysia") , 
                   CO2.emissions ~ Total.Forest + Oil.Palm + Year)
summary(CO2_mly)
```

#Carbon Emissions Explanation
> CO2.emissions which is our variable of interest in the release of carbon emissions through land use change and forestry. We fee this is best suited for our analysis as the variable carbon emissions from agriculture explains emissions from livestock. To begin the analysis, we first use the AIC to gauge which variables best impact carbon dioxide emissions. The lowest AIC from this stepwise regression suggests that we use Total Forests (inluding natural and planted growth), Oil Palm cultivation and year. This suggests that planted forests too impact carbon emissions which makes for a positive argument in support of their credibility. 
>The MLR for Indonesia reports a very low R squared(3.96%). We see a negative correlation between carbon emissions and total forests (-0.21) and year(-17.77) with the former having a smaller impact than the later. As expected, carbon emissions increase with oil palm cultivation (0.013). 
> The MLR for Malaysia reports an R squared of 60.83%. We see a large negative coorelation between carbon emissions and total forests (-199.5). Surprisingly, oil palm cultivation too has an inverse relationship with carbon emissions in Malaysia (-0.04) while it is a relatively smaller number. 

```{r}
dataset_ind <- dataset %>%
  subset(Entity == "Indonesia") %>%
  mutate(Date = paste(Year, "-", 01, "-", 01))
dataset_ind$Date <- ymd(dataset_ind$Date)

#TIME SERIES ANALYSIS 
##create ts object
natural_forest_ind_ts <- ts(dataset_ind$Natural.Forest, start = c(1990,01,01), frequency = 1)
#run a mann-kendall test on this data
mk.test(natural_forest_ind_ts)
##plot the data
natural_forest_ind_plot <-
ggplot(dataset_ind, aes(x = Date, y = Natural.Forest)) +
  geom_point() +
  geom_line() +
  ylab("Natural Forest (million ha)") +
  xlab("Year") +
  geom_smooth(method = lm ) +
  labs(title = "Natural Forest in Indonesia") +
  mytheme
print(natural_forest_ind_plot)


##repeat the same for Malaysia
dataset_mly <- dataset %>%
  subset(Entity == "Malaysia") %>%
  mutate(Date = paste(Year, "-", 01, "-", 01))
dataset_mly$Date <- ymd(dataset_mly$Date)
##create ts object
natural_forest_mly_ts <- ts(dataset_mly$Natural.Forest, start = c(1990,01,01), frequency = 1)
#run a mann-kendall test on this data
mk.test(natural_forest_mly_ts)
##plot the data
natural_forest_mly_plot <-
ggplot(dataset_mly, aes(x = Date, y = Natural.Forest)) +
  geom_point() +
  geom_line() +
  ylab("Natural Forest (million ha)") +
  xlab("Year") +
  geom_smooth(method = lm ) +
  labs(title = "Natural Forest in Malaysia") +
  mytheme
print(natural_forest_mly_plot)

```

\newpage

# Summary and Conclusions


\newpage

# References
World Wildlife Fund. (n.d.). Palm oil. World Wildlife Fund. https://www.worldwildlife.org/industries/palm-oil

Ritchie, H. (2021). Palm oil. Our World in Data. https://ourworldindata.org/palm-oil

Ritchie, H, Rosado, P and Roser, M (2020). Emissions by sector: where do GHGs come from?.Our World in Data. https://ourworldindata.org/emissions-by-sector

Ritchie, H. (2021). Forest Area. Our World in Data. https://ourworldindata.org/forest-area 


